Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-02 manuscript.tex
Outline of results section
Need to review and join sentences.
Figures are missing and have only placeholders

$D$ indirectly controls both the accuracy of the integration
and the computation time.

@@

\citet{Li2011} provide a few examples of values for this ratio in the case of
the vertical component of the gravitational attraction.

In this section, we investigate
the relationship between
values of the distance-size ratio
and the computation error.

We perform the error analysis for the gravitational potential,
acceleration, and gradient tensor components
to evaluate if the same value of $D$ yields compatible error levels
for different fields.

We establish a homogeneous spherical shell as a reference against which to
compare the computed tesseroid fields.

The shell has analytical solutions for the gravitational potential,
attraction, and gradient tensor along the polar axis
\citep{Grombein2013}.

We used a spherical shell with a thickness of 1 km,
density of $2670\ kg.m^{-3}$,
bottom at height 0 km above the reference sphere,
and top at 1 km height.

We discretized the spherical shell along the horizontal dimensions into
a regular mesh of tesseroids.

If we calculate the tesseroid fields at a single point, that point might not
coincide with the point of largest error.

Figure~\ref{fig:glqerrorsample} shows that the largest errors are concentrated
on top of the tesseroid.

Performing the error analysis on calculations at a single point will not be
representative of the true error.

@@

We calculate the effect of the tesseroid model of the spherical shell
on a regular grid.

The effect of the shell will be same along the entire grid.

We compute the differences between the effects of the shell and the tesseroid
model on the grid and keep only the largest error.

The grid is at a constant height of 2 km (1 km above the shell).

The grid is located on top of a tesseroid located at the North pole.

The computation point can be considered at any point with equal radial
coordinate located outside the shell because of the symmetry of the shell.

We calculate the difference between the shell values and the values computed
from tesseroids.

We keep only the largest difference.

Calculating on a grid above a tesseroid is important because
the largest errors in the GLQ integration occur directly on top of the
tesseroid but not exactly at it's center, as can be seen in
Figure~\ref{fig:glqerrorsample}.

The difference between shell and tesseroid values would be underestimated
if the computation was performed only at a single point on the polar axis.

We repeated the tesseroid calculations with values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.

Figure~\ref{fig:tess_vs_shell} shows the maximum difference between the
shell and tesseroid fields as a function of $D$.

@@

The error versus $D$ curves allow us to choose a value of $D$ that
results in an error below a chosen maximum.

We chose 0.1\% as the maximum allowed error.

We performed the computations again for three different scenarios: at 260 km
height (the mean orbit height of the GOCE satellite), on top of a tesseroid at
the equator, and for a $30^\circ \times 30^\circ \times 1\ km$ tesseroid model
of the shell.

We performed these computations to investigate if the GLQ error decays the same
way under different scenarios.

The error curves for the gravitational potential show increased errors at the
equator and for a $30^\circ$ tesseroid.

For $D=1$ all errors are smaller than 0.1\%,
even in the extreme case of a $30^\circ$ tesseroid with computation points at a
height of 1 km above the tesseroid.

$D=1.5$ is sufficient to achieve errors below 0.1\% in $g_z$.

$D=8$ is required to achieve errors below 0.1\% in
the gravity gradient components.

The error curves for the computation at 260 km height shows that
the error is smaller at higher altitudes for all fields.

Computation at 260 km height does not produce errors above the 0.1\% threshold
for the potential and $g_z$.

The computation height changes the value of the distance-size ratio required
for the gravity gradient components.

We computed the error curves for $g_{zz}$ at different heights
(Figure~\ref{fig:gzz_error_height}).

The distance-size ratio can be reduced to 4 if the computation is done at 50 km
height.

One can use a distance-size ratio of 2.5 for computations at 260 km height.

These reductions in the distance-size ratio can reduce the computation time
required to achieve the desired.

The software \textit{Tesseroids} includes default values of the distance-size
ratio.

We used the values required to achieve 0.1\% accuracy: 1 for the potential,
1.5 for the gravitational attraction, and 8 for the gravity gradient
components.

The value of $D=8$ for the gravity gradients is conservative.

We chose it to ensure the accuracy of the computation in most cases.

The experienced user of the software can set smaller values using command-line
options.

The smaller values result in less computation time.

>>>>

@@

Thus, $D$ indirectly controls both the accuracy of the integration
and the computation time.

In this section, we investigate the relationship between
values of the distance-size ratio and the integration error.

We perform the analysis for the gravitational potential,
acceleration, and gradient tensor components
to evaluate if the same value of $D$ yields compatible error levels
for different fields.

The reference against which we compare the computed tesseroid fields
is a homogeneous spherical shell.

The shell has analytical solutions along the polar axis
\citep{Grombein2013}
and can be perfectly discretized into tesseroids.

We chose a spherical shell with a thickness of 1 km,
density of $2670\ kg.m^{-3}$,
bottom at height 0 km above the reference sphere,
and top at 1 km height.

We produced tesseroid models of the shell by discretizing it along the
horizontal dimensions into a regular mesh.

Figure~\ref{fig:glqerrorsample} shows that the largest errors are spread over
on top of the tesseroid.

Thus, calculating the tesseroid fields at a single point might not
capture the point of largest error.

Instead, we calculate the effect of the tesseroid model on a regular grid
of $10 \times 10$ points.

@@

Therefore, the effect of the shell will be same along the entire grid.

We compute the differences between the effects of the shell and the tesseroid
model on the grid and consider only the largest error.

By placing the grid on top of a particular tesseroid, we increase the chances
of capturing the true largest integration error.

We calculate the errors for values of the distance-size ratio
$D$ varying from 0 (i.e., no divisions) to 10 in 0.5 intervals.

We repeated the analysis in four different numerical experiments,
each with computation grids at different locations and
different tesseroid model sizes.

Table~\ref{tbl:experiment} describes the different numerical
experiments and the corresponding parameters of the computation grid and
tesseroid model.

Figure~\ref{fig:dist-size-curves} shows
the maximum difference between the shell and tesseroid fields
as a function of $D$ for the four numerical experiments.

@@

We established a maximum tolerated error of $0.1\%$, represented by the
horizontal solid lines in Figure~\ref{fig:dist-size-curves}.

For the potential $V$, a distance-size ratio $D=1$ guarantees that the curves
for all experiments are below the $0.1\%$ error threshold.

For $g_z$, the same is achieved with $D=1.5$.

Conversely, $g_{zz}$ requires a value of $D=8$ to achieve an error level of
$0.1\%$.

For a computation height of 260 km, the error curve for $G_{zz}$
intercepts the error threshold line at $D=2.5$.

This behavior suggests that the error curves for $g_{zz}$ might depend on the
computation height.

To test this hypothesis, we computed the error curves for $g_{zz}$ at
heights 2, 10, 50, 150, and 260 km.

Figure~\ref{fig:gzz-with-height} shows the results for $g_{zz}$ at varying
computation heights.

Notice that the distance-size ratio needed to achieve $0.1\%$ accuracy
decreases as the computation height increases.

For example, computation at 260 km height requires $D=2.5$ whereas at 10 km
height a value of $D=5.5$ is required to achieve the same accuracy.

One can take advantage of this behavior to reduce the distance-size ratio for
computations of the gravity gradient tensor at high altitudes,
saving computation time.

We have implemented the values of the distance-size ratio producing $0.1\%$
accuracy as defaults for the software \textit{Tesseroids}.

We chose the conservative value of $D=8$ for the gravity gradient components
as a fail-safe alternative.

Users can control the value of $D$ used in the computations through
command-line arguments to achieve greater performance as a trade-off for
accuracy.

\begin{table*}
\centering
\begin{tabular}{cccc}
    \hline
    ~             & Grid location & Grid height & Tesseroid size \\
    \hline
    Experiment 1  & 89N--90N/0E--1E & 2 km & $1^\circ \times 1^\circ$ \\
    Experiment 2  & 0N--1N/0E--1E & 2 km & $1^\circ \times 1^\circ$ \\
    Experiment 3  & 89N--90N/0E--1E & 260 km & $1^\circ \times 1^\circ$ \\
    Experiment 4  & 60N--90N/0E--30E & 2 km & $30^\circ \times 30^\circ$ \\
    \hline
\end{tabular}
\caption{Parameters of the numerical experiments.}
\label{tbl:experiment}
\end{table*}
